---
title: "288 Project_Decision Tree"
author: "Kathy Yu Hsin Lee"
date: "6/5/2021"
output: html_document
---
**Packages **
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)
library(zoo)
library(lubridate)
library(stringr)
library(fpp2)
library(leaps)
library(tree)
```

## Data Cleaning ##
**Data Cleaing**
```{r}
# Clean Data
setwd("/Users/KathyLee/UCI MSBA/Course Material/05 Spring Quarter/BANA288 Predictive Analytics/Project")
dat <- read.csv("12-2019.csv") # import august flight

# Delete unused columns, remove national and security
flight <- dat[-c(2,5,7:9,12,13,15,16,17,22:24,30)]

# transform delay reason into binary variables
flight$delay_carrier <- ifelse(flight$delay_carrier>15, 1, 0)
flight$delay_weather <- ifelse(flight$delay_weather>15, 1, 0)
flight$delay_late_aircarft_arrival <- ifelse(flight$delay_late_aircarft_arrival>15, 1, 0)

# change schedule departure and arrival dt to numeric
dp_time <- strftime(flight$scheduled_departure_dt, format="%H:%M")
ar_time <- strftime(flight$scheduled_arrival_dt, format="%H:%M")
flight$scheduled_departure_dt <- as.numeric(sub("^(\\d+):(\\d+).*", "\\1.\\2", dp_time))
flight$scheduled_arrival_dt <- as.numeric(sub("^(\\d+):(\\d+).*", "\\1.\\2", ar_time))

```

**Select Only Top 30 Airport Data **
```{r}
# choose top 30 airports
top <- c('MDW','BNA','PDX','TPA','IAD','SAN','MIA','BWI','FLL','SLC',  
         'JFK','DCA','BOS','PHL','MSP','MCO','EWR','LAS','LGA','SFO',
         'DTW','IAH','PHX','SEA','LAX','CLT','DEN','DFW','ATL','ORD')

top30 <- flight %>% filter(origin_airport %in% top & destination_airport %in% top)

# remove na
top30 <- na.omit(top30)

table(top30$delay_weather)
table(top30$delay_carrier)
table(top30$delay_late_aircarft_arrival)
```

# Decision Tree #

```{r}
library(rpart)
library(rpart.plot)
library(rattle)
```

## Carrier Delay##
**Data Preparation**
```{r}
# subset delay and ontime for carrier delay
delay <- top30 %>% filter(delay_carrier==1)
ontime <- top30 %>% filter(delay_carrier==0)

set.seed(123456)

train.delay <- sample(nrow(delay),5000)
train.ontime <- sample(nrow(ontime),5000)
train1 <- rbind(delay[train.delay,], ontime[train.ontime,])

test.delay <- sample(nrow(delay[-train.delay]), 1250)
test.ontime <- sample(nrow(ontime[-train.ontime]), 1250)
test1 <- rbind(delay[test.delay,], ontime[test.ontime,])

# Prepare Data for decision tree
#  Set up the Y variable as a factor 
train1$delay_carrier <- as.factor(train1$delay_carrier)
test1$delay_carrier <- as.factor(test1$delay_carrier)

#  Set up indicators for all the factors
train1$carrier_code <- as.factor(train1$carrier_code)
train1$origin_airport <- as.factor(train1$origin_airport)
train1$destination_airport <- as.factor(train1$destination_airport)
train1$day <- as.factor(train1$day)
train1$weekday <- as.factor(train1$weekday)

test1$carrier_code <- as.factor(test1$carrier_code)
test1$origin_airport <- as.factor(test1$origin_airport)
test1$destination_airport <- as.factor(test1$destination_airport)
test1$day <- as.factor(test1$day)
test1$weekday <- as.factor(test1$weekday)

# Remove other delay column
train1 <- train1[-c(6,7)]
test1 <- test1[-c(6,7)]
```

**Build a tree**
```{r}
#  Build a tree on the training data
tree.train <- rpart(delay_carrier ~ ., data = train1, method='class')
summary(tree.train)

# Plot the tree
fancyRpartPlot(tree.train, main = "Classification Tree on Carrier Delay", caption = NULL)
rpart.plot(tree.train, extra = 2)

```

**Important Variables**
```{r}
tree.train$variable.importance
printcp(tree.train)
```

**Check performance on train1 and test1**
```{r}
#  Predict on train1
tree.train.pred <- predict(tree.train, train1)
tree.train.pred.cl <- 
  ifelse(tree.train.pred[,2] > 0.5, 1, 0)
tab.train <- table(train1$delay_carrier, tree.train.pred.cl, dnn = c("Actual", "Predicted"))
tab.train
mean(train1$delay_carrier != tree.train.pred.cl)

# Predict on test1
tree.test.pred <- predict(tree.train, test1)
tree.test.pred[1:10,]
tree.test.pred.cl <- ifelse(tree.test.pred[,2] > 0.5, 1, 0)
tab.test <- table(test1$delay_carrier, tree.test.pred.cl, dnn = c("Actual", "Predicted"))
tab.test
err.tree.test <- mean(test1$delay_carrier != tree.test.pred.cl)
err.tree.test
```


## Late Aircraft Delay ##
**Data Preparation**
```{r}
# subset delay and ontime for late aircraft delay
delay <- top30 %>% filter(delay_late_aircarft_arrival==1)
ontime <- top30 %>% filter(delay_late_aircarft_arrival==0)

set.seed(123456)

train.delay <- sample(nrow(delay),5000)
train.ontime <- sample(nrow(ontime),5000)
train2 <- rbind(delay[train.delay,], ontime[train.ontime,])

test.delay <- sample(nrow(delay[-train.delay]), 1250)
test.ontime <- sample(nrow(ontime[-train.ontime]), 1250)
test2 <- rbind(delay[test.delay,], ontime[test.ontime,])

# Prepare Data for decision tree
#  Set up the Y variable as a factor 
train2$delay_late_aircarft_arrival <- as.factor(train2$delay_late_aircarft_arrival)
test2$delay_late_aircarft_arrival <- as.factor(test2$delay_late_aircarft_arrival)

#  Set up indicators for all the factors
train2$carrier_code <- as.factor(train2$carrier_code)
train2$origin_airport <- as.factor(train2$origin_airport)
train2$destination_airport <- as.factor(train2$destination_airport)
train2$day <- as.factor(train2$day)
train2$weekday <- as.factor(train2$weekday)

test2$carrier_code <- as.factor(test2$carrier_code)
test2$origin_airport <- as.factor(test2$origin_airport)
test2$destination_airport <- as.factor(test2$destination_airport)
test2$day <- as.factor(test2$day)
test2$weekday <- as.factor(test2$weekday)

# Remove other delay column
train2 <- train2[-c(5,6)]
test2 <- test2[-c(5,6)]
```

**Build a tree**
```{r}
#  Build a tree on the training data
tree.train2 <- rpart(delay_late_aircarft_arrival ~ ., data = train2, method='class')
summary(tree.train2)

# Plot the tree
fancyRpartPlot(tree.train2, caption = NULL, main = "Classification Tree on Late Aircraft Delay")
#rpart.plot(tree.train2)

```

**Important Variables**
```{r}
tree.train2$variable.importance
printcp(tree.train2)
```

**Check performance on train2 and test2**
```{r}
#  Predict on train2
tree.train.pred2 <- predict(tree.train2, train2)
tree.train.pred.c2 <- 
  ifelse(tree.train.pred2[,2] > 0.5, 1, 0)
tab.train2 <- table(train2$delay_late_aircarft_arrival, tree.train.pred.c2, dnn = c("Actual", "Predicted"))
tab.train2
mean(train2$delay_late_aircarft_arrival != tree.train.pred.c2)

# Predict on test2
tree.test.pred2 <- predict(tree.train2, test2)
tree.test.pred2[1:10,]
tree.test.pred.c2 <- ifelse(tree.test.pred2[,2] > 0.5, 1, 0)
tab.test2 <- table(test2$delay_late_aircarft_arrival, tree.test.pred.c2, dnn = c("Actual", "Predicted"))
tab.test2
err.tree.test2 <- mean(test2$delay_late_aircarft_arrival != tree.test.pred.c2)
err.tree.test2
```


## Weather Delay ##
**Data Preparation**
```{r}
# subset delay and ontime for weather delay
delay <- top30 %>% filter(delay_weather==1)
ontime <- top30 %>% filter(delay_weather==0)

set.seed(123456)

train.delay <- sample(nrow(delay),914)
train.ontime <- sample(nrow(ontime),914)
train3 <- rbind(delay[train.delay,], ontime[train.ontime,])

test.delay <- sample(nrow(delay[-train.delay]), 229)
test.ontime <- sample(nrow(ontime[-train.ontime]), 229)
test3 <- rbind(delay[test.delay,], ontime[test.ontime,])

# Prepare Data for decision tree
#  Set up the Y variable as a factor 
train3$delay_weather <- as.factor(train3$delay_weather)
test3$delay_weather <- as.factor(test3$delay_weather)

#  Set up indicators for all the factors
train3$carrier_code <- as.factor(train3$carrier_code)
train3$origin_airport <- as.factor(train3$origin_airport)
train3$destination_airport <- as.factor(train3$destination_airport)
train3$day <- as.factor(train3$day)
train3$weekday <- as.factor(train3$weekday)

test3$carrier_code <- as.factor(test3$carrier_code)
test3$origin_airport <- as.factor(test3$origin_airport)
test3$destination_airport <- as.factor(test3$destination_airport)
test3$day <- as.factor(test3$day)
test3$weekday <- as.factor(test3$weekday)

# Remove other delay column
train3 <- train3[-c(5,7)]
test3 <- test3[-c(5,7)]
```

**Build a tree**
```{r}
#  Build a tree on the training data
tree.train3 <- rpart(delay_weather ~ ., data = train3, method='class')
summary(tree.train3)

# Plot the tree
fancyRpartPlot(tree.train3, caption = NULL, main = "Classification Tree on Weather Delay")
#rpart.plot(tree.train3)

```

**Important Variables**
```{r}
tree.train3$variable.importance
printcp(tree.train3)
```

**Check performance on train3 and test3**
```{r}
#  Predict on train3
tree.train.pred3 <- predict(tree.train3, train3)
tree.train.pred.c3 <- 
  ifelse(tree.train.pred3[,2] > 0.5, 1, 0)
tab.train3 <- table(train3$delay_weather, tree.train.pred.c3, dnn = c("Actual", "Predicted"))
tab.train3
mean(train3$delay_weather != tree.train.pred.c3)

# Predict on test3
tree.test.pred3 <- predict(tree.train3, test3)
tree.test.pred3[1:10,]
tree.test.pred.c3 <- ifelse(tree.test.pred3[,2] > 0.5, 1, 0)
tab.test3 <- table(test3$delay_weather, tree.test.pred.c3, dnn = c("Actual", "Predicted"))
tab.test3
err.tree.test3 <- mean(test3$delay_weather != tree.test.pred.c3)
err.tree.test3
```


## Boosting ##
```{r}
library(gbm)
```

**Change the Y variable back to quantitative 0/1**
```{r}
train1$delay_carrier <- as.numeric(train1$delay_carrier) - 1
test1$delay_carrier <- as.numeric(test1$delay_carrier) - 1

train2$delay_late_aircarft_arrival <- as.numeric(train2$delay_late_aircarft_arrival) - 1
test2$delay_late_aircarft_arrival <- as.numeric(test2$delay_late_aircarft_arrival) - 1

train3$delay_weather <- as.numeric(train3$delay_weather) - 1
test3$delay_weather <- as.numeric(test3$delay_weather) - 1
```

**Carrier Delay**
```{r}
# Carrier Delay
boost.train <- gbm(delay_carrier ~ ., data = train1, 
                   distribution = "bernoulli", n.trees = 10000, 
                   interaction.depth = 7, shrinkage = 0.005)
summary(boost.train)

#  Top 10 variables 
summary(boost.train)[1:10,]

# Carrier Delay Prediction
yhat.boost <- predict(boost.train, test1, n.trees= 10000, type = "response")
yhat.boost
yhat.boost.cl <- ifelse(yhat.boost > 0.5, 1, 0)
tab.boost <- table(test1$delay_carrier, yhat.boost.cl)
tab.boost
err.boost <- mean(test1$delay_carrier != yhat.boost.cl)
err.boost
```

**Late Aircraft Arrival Delay **
```{r}
# Late Aircraft Arrival Delay
boost.train2 <- gbm(delay_late_aircarft_arrival ~ ., data = train2, 
                   distribution = "bernoulli", n.trees = 10000, 
                   interaction.depth = 7, shrinkage = 0.005)
summary(boost.train2)

#  Top 10 variables 
summary(boost.train2)[1:10,]

# Late Aircraft Arrival Delay Prediction
yhat.boost2 <- predict(boost.train2, test2, n.trees= 10000, type = "response")
yhat.boost2
yhat.boost.c2 <- ifelse(yhat.boost2 > 0.5, 1, 0)
tab.boost2 <- table(test2$delay_late_aircarft_arrival, yhat.boost.c2)
tab.boost2
err.boost2 <- mean(test2$delay_late_aircarft_arrival != yhat.boost.c2)
err.boost2

```

**Weather Delay**
```{r}
# Weather Delay
boost.train3 <- gbm(delay_weather ~ ., data = train3, 
                   distribution = "bernoulli", n.trees = 10000, 
                   interaction.depth = 7, shrinkage = 0.005)
summary(boost.train3)

#  Top 10 variables
summary(boost.train3)[1:10,]

# Weather Delay Prdiction
yhat.boost3 <- predict(boost.train3, test3, n.trees= 10000, type = "response")
yhat.boost3
yhat.boost.c3 <- ifelse(yhat.boost3 > 0.5, 1, 0)
tab.boost3 <- table(test3$delay_weather, yhat.boost.c3)
tab.boost3
err.boost3 <- mean(test3$delay_weather != yhat.boost.c3)
err.boost3
```

